\bf Model Visualization Techniques for a Social Network Model

### 2017-08-15 paper.Rnw
add results from senate model fitting; fiddle with the axes font & labels in plots

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.8\\linewidth', par=TRUE, cache=TRUE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
@

@@

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists} i\to j \\
                                   0 & otherwise
  \end{cases}
\]
In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .05, directed = T) +
  theme_net() + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) +
  xlab("from (ego)") +
  ylab("to (alter)")
@
\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data. \st{The p-value shown on the x-axis is .... }", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
p1 <- plot(gof1)
p2 <- plot(gof2)
gridExtra::grid.arrange(p1, p2)
@

@@

To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable. The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}. 
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = x, y=y)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "i", y = "j") + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 3, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
  theme_net() +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@

@@

set.seed(34569234)
create_net_animate(static_pall) + 
  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
  theme(panel.background = element_rect(color = 'black'))
@

@@

<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id. On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8>>=
# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
# reinstall new dplyr
numordersf <- data.frame(fd2.w1) %>% 
  mutate(from = rownames(fd2.w1)) %>% 
  gather(key = to ,value = value, -from)
pcaordersf <- numordersf

@@

p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
  scale_x_discrete(position = "top") + 
  coord_fixed()

@@

p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
  scale_x_discrete(position = "top") + 
  coord_fixed()

@@

ggplot(data = changes3) + 
  geom_segment(
    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type)) + 
  geom_point(aes(x = alter, y = empprob), shape = 4, size = 1, alpha = .5) + 
  labs(x = "alter node", y = "probability of alter node") + 
  coord_flip() + 
  facet_wrap(~ego, labeller = "label_both") + 
  theme_bw()
# 
# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()

@@

Another way to view the transition probabilities is through the adjacency matrix visualization. In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse, showing the lack of coverage in the model.  Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains. XXX Check this again.... XXX
<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations. The acting node is on the x-axis, and the receiving node is on the y-axis. Missing squares represent ties that did not occur.'>>=
ansnullchains %>% 
  filter(rep == 1, period == 1) %>% 
  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
         alter = paste0("V",as.numeric(as.character(alter))+1), 
         prob = exp(logAlterProb)) -> probsDat

ggplot(data = probsDat, 
       aes(x = factor(ego, levels = order1), 
             y = factor(alter, levels = rev(order1)), 
             fill = prob)) + 
  geom_tile(color = 'black') + 
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  theme(panel.background = element_rect(fill = 'white')) + 
  labs(x = "ego", y = "alter")+
  scale_x_discrete(position = 'top') + 
  coord_fixed()
@

@@

<<allsteps, fig.cap="Three repetitions of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data.">>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to), 
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>% 
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>% 
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

p1 <- ggplot(data = msall1_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,81.5)) + 
  scale_fill_continuous(low = "red", high = "blue") + 
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) +
  labs(x = "Microstep")

@@

p2 <- ggplot(data = msall2_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,49.5)) + 
  scale_fill_continuous(low = "red", high = "blue") +
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) #+
  #labs(x = "Microstep", y = "Edges (most frequent at top)")

@@

p3 <- ggplot(data = msall3_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,106.5)) + 
  scale_fill_continuous(low = "red", high = "blue") +
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) #+
  #labs(x = "Microstep", y = "Edges (most frequent at top)")

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = t1$color,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
           singletons = T) +
  theme_net() + labs(title = "Summary Network: M1") + 
  theme(plot.background = element_rect(color = 'black'),
        plot.title = element_text(size = rel(.5)))

p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.1, "inches")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
           singletons = T) +
  theme_net() + labs(title = "Wave 1") + 
    theme(plot.background = element_rect(color = 'black'),
        plot.title = element_text(size = rel(.5)))
p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02,
           singletons = T) +
  theme_net() + labs(title = "Wave 2")  + 
  theme(plot.background = element_rect(color = 'black'),
        plot.title = element_text(size = rel(.5))) 
grid.arrange(p2, p1, p3, nrow = 1)
@

@@

\begin{figure}\label{fig:structures}
\begin{subfigure}[t]{.45\textwidth}
\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary. The group of the actors is represented by the shape of the node.}
<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center'>>=
jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))

>>>>

<<setup, echo = FALSE, message = FALSE, warning = FALSE>>=
options(replace.assign=TRUE,width=70, digits=3)
require(knitr)
opts_chunk$set(fig.path='figure/', cache.path='cache/', fig.align='center', fig.pos='h', out.width='.99\\linewidth', par=TRUE, cache=FALSE, concordance=TRUE, autodep=TRUE, message=F, warning=F, echo = FALSE)#, root.dir = "~/Desktop/Dissertation/SAOM-removing-blindfold/")
@

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)

ThemeNoNet <- theme_bw() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'bold', angle = 0), 
                  axis.title.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0),
                  axis.title.y = element_text(size = rel(2),
                                            face = 'plain', angle = 90),
                  axis.text.x.top = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.x.bottom = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0), 
                  axis.text.y.left = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  axis.text.y.right = element_text(size = rel(1.5),
                                            face = 'plain', angle = 0),
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 90,margin = margin(t = 3, r = 3, b = 3, l = 3, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white")
                    )
ThemeNet <- theme_net() %+replace% 
            theme(plot.title = element_text(size = rel(2),
                                            face = 'bold', angle = 0), 
                  strip.text.x = element_text(size = rel(2),
                                            face = 'plain', angle = 0, 
                                            margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.text.y =element_text(size = rel(2),
                                            face = 'plain', angle = 0,margin = margin(t = 3, r = 0, b = 3, l = 0, unit = "pt")),
                  strip.background = element_rect(colour = "black", fill = "white"))
@

@@

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists} i\to j \\
                                   0 & otherwise
  \end{cases}
\]
In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 4, arrowgap = .05, directed = T) +
  ThemeNet + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  ThemeNoNet + 
  theme(panel.grid = element_blank(), aspect.ratio = 1) +
  labs(title = "Adjacency Matrix Visualization") + 
  xlab("from (ego)") +
  ylab("to (alter)")
@
\hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data. \\st{The p-value shown on the x-axis is .... }", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
p1 <- plot(gof1)
p2 <- plot(gof2)
gridExtra::grid.arrange(p1, p2)
@

@@

To guide our visual exploration of stochastic actor-oriented models, we used a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy, with changes that are more noticeable. The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The actor covariates and networks are visualized in Figures~\ref{fig:adjmatviz} and~\ref{fig:alldataviz}. 
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", fig.width=8, results='hide',fig.show='asis', fig.width=8, fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = x, y=y)) + 
  geom_tile(aes(fill = as.factor(value))) + 
  scale_fill_manual("i likes j", labels=c("no", "yes"), 
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymin = 19.5, ymax = 35.5, fill =NA)) + 
  facet_wrap(~wave) + theme_bw() +
  labs(x = "i", y = "j") + 
  ThemeNoNet + 
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank()) 
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@

@@

set.seed(34569234)
create_net_animate(static_pall) + 
  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
  ThemeNet + 
  theme(panel.background = element_rect(color = 'black'))
@

@@

<<adjmatorder, message = FALSE, warning = FALSE, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id. On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8, fig.height=3>>=
# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
# reinstall new dplyr
numordersf <- data.frame(fd2.w1) %>% 
  mutate(from = rownames(fd2.w1)) %>% 
  gather(key = to ,value = value, -from)
pcaordersf <- numordersf

@@

p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top", labels = as.character(1:16)) +
  scale_y_discrete(labels = rev(as.character(1:16)))  + 
  ThemeNoNet + 
  theme(legend.position = 'none') + 
  coord_fixed()

@@

p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) + 
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) + 
  ThemeNoNet + 
  theme(legend.position = 'none') +
  coord_fixed()

@@

ggplot(data = changes3) + 
  geom_segment(
    aes(x = alter , xend = alter, y = 0, yend = alterProb, color=type), size = 2) + 
  geom_point(aes(x = alter, y = empprob), shape = 4, size = 2, alpha = .5) + 
  labs(x = "alter node", y = "probability of alter node") + 
  coord_flip() + 
  facet_wrap(~ego, labeller = "label_both") + 
  ThemeNoNet
# 
# cumprobs %>% ggplot() + geom_histogram(aes(x = alter, weight = count), color = 'black', fill = 'white', binwidth = 1) + scale_x_continuous(labels = paste0("V",1:16), breaks = 0:15) + coord_flip() + facet_wrap(~ego, labeller = "label_both") + theme_gray()

@@

Another way to view the transition probabilities is through the adjacency matrix visualization. In Figure~\ref{fig:heatmap}, we build on the concept of the ordered adjacency matrix of Figure~\ref{fig:adjmatorder}. This heatmap shows the transition probabilities of all ties that are changed in the first microstep of the 1,000 simulations. The heatmap is noticeably sparse, showing the lack of coverage in the model.  Of the 256 possible steps, only 81, or about 32\%, are taken in the 1,000 simulated chains. XXX Check this again.... XXX
<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations. The acting node is on the x-axis, and the receiving node is on the y-axis. Missing squares represent ties that did not occur.', fig.height=3>>=
ansnullchains %>% 
  filter(rep == 1, period == 1) %>% 
  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
         alter = paste0("V",as.numeric(as.character(alter))+1), 
         prob = exp(logAlterProb)) -> probsDat

ggplot(data = probsDat, 
       aes(x = factor(ego, levels = order1), 
             y = factor(alter, levels = rev(order1)), 
             fill = prob)) + 
  geom_tile(color = 'black') + 
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  labs(x = "ego", y = "alter")+
  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + theme(panel.grid = element_blank()) + 
  coord_fixed()
@

@@

<<allsteps, fig.cap="Three simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.">>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to), 
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>% 
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>% 
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

p1 <- ggplot(data = msall1_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,81.5)) + 
  scale_fill_continuous(low = "red", high = "blue") + 
  ThemeNoNet + 
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) +
  labs(x = "Microstep")

@@

p2 <- ggplot(data = msall2_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,49.5)) + 
  scale_fill_continuous(low = "red", high = "blue") +
  ThemeNoNet + 
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) #+
  #labs(x = "Microstep", y = "Edges (most frequent at top)")

@@

p3 <- ggplot(data = msall3_df2) + 
  geom_tile(aes(x = ms, y = -eid, fill = ms)) + 
  geom_vline(xintercept = c(0.5,106.5)) + 
  scale_fill_continuous(low = "red", high = "blue") +
  ThemeNoNet + 
  theme(legend.position = 'none', axis.ticks.y = element_blank(),
        axis.text.y = element_blank(), axis.title = element_blank()) #+
  #labs(x = "Microstep", y = "Edges (most frequent at top)")

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1$color,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Summary: M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))

p2 <- ggplot(data = filter(combinedavgs, cat == "First Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
   labs(title = "Wave 1") + ThemeNet + 
    theme(plot.background = element_rect(color = 'black'))
p3 <- ggplot(data = filter(combinedavgs, cat == "Second Observation")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
  labs(title = "Wave 2")  +  ThemeNet + 
  theme(plot.background = element_rect(color = 'black')) 
grid.arrange(p2, p1, p3, nrow = 1)
@

@@

\begin{figure}\label{fig:structures}
\begin{subfigure}[t]{.45\textwidth}
\caption{\label{fig:jtt}Realization of a jumping transitive triplet, where $i$ is the focal actor, $j$ is the target actor, and $h$ is the intermediary. The group of the actors is represented by the shape of the node.}
<<jtt, echo=FALSE, fig.width=2, fig.height=2, message = FALSE, warning = FALSE, fig.align='center',out.width='.49\\linewidth'>>=
jTTe <- data.frame(from = c('i', 'i', 'h'), to = c('h', 'j', 'j'))
jTTn <- data.frame(id = letters[8:10], group = c(1,1,2))
